A new metric for automatic discovery of periodic patterns in time series

### 2021-03-03 paper/main.Rmd
heatmap changed and discussion added

---
title: A new metric for automatic discovery of periodic patterns in time series
authors:
- name: Sayani Gupta
  affiliation: Department of Econometrics and Business Statistics, Monash University
  email: Sayani.Gupta@monash.edu

@@

The smart meter data set for eight households in Melbourne has been utilized to see the use of the $wpd_{norm}$ proposed in the paper. The data has been cleaned to be a `tsibble` (@wang2019tsibble) containing half-hourly electricity consumption from Jul-2019 to Dec-2019 for each of the households, which is procured by them by downloading their data from the energy supplier/retailer. Demand data for these households are shown in a linear time scale in Figure \ref{fig:linear-scale-8}. It is evident from the range of the demand data that these households vary in consumption levels as well as in their temporal patterns. In the left panel of Figure \ref{fig:linear-scale-8} (a), the linear representation of the entire time period is shown, whereas in the right panel (b) a particular month is shown and furthermore a week has been highlighted to inspect if there is any daily or weekly periodic patterns in their behavior that is reflected when we zoom into the the linear representation of the time series. We start the analysis by asking if the ranking of the harmonies make sense for the households, then compare households to get more insights of what these rankings imply and if they could be used to remove some non-interesting harmonies. Furthermore, we see if the display of the significant harmonies could be validated by zooming in the linear representation of the time series.

@@

@@

_Choosing and Ranking harmonies for all households_

@@

```{r gravitas-plot-8, message = FALSE, warning=FALSE, fig.cap="Distribution of energy demand shown for household id 1 across hod in x-axis and wd-wnd in facets in a) and just the reverse in b). In c), distribution of energy demand for household id:4 shown across hod and wd-wnd. It can be seen that the differences in distributions are more apparent when viewed in a) as compared to b). It seems like there is more difference in the distributions of hod for b) compared to c). This also confers with the value of the normalised measure shown in Figure 11."}

@@

From Figure \ref{fig:gravitas-plot-8-id5}, it could further be observed that id5 has only one significant harmony $(hod, dow)$. Apparently, $(hod, wnd/wday)$ which is an important harmony for most households is not important for this one.

_Comparing similar households and spotting anomalous ones_

```{r gravitas-plot-8-id5, message = FALSE, warning=FALSE, fig.cap="something"}

@@

_Validating patterns from linear display_

```{r raw-nqt-8, fig.cap="The raw density of the half-hourly demand for the eight households in Panel a. Panel b shows the normal-score-transform half-hourly demand for the same households which has resulted in more symmetric distribution of half-hourly demand."}
elec_raw <- elec %>% 
  ggplot() +
  geom_density(aes(x = kwh)) +
  facet_wrap(~id, scales = "free_y", ncol = 1) +
  xlab("demand (in kwh)")

elec_nqt <- elec %>% 
  dplyr::mutate(kwh_transformed = stats::qqnorm(kwh, plot.it = FALSE)$x) %>% 
  ggplot() +
  geom_density(aes(x = kwh_transformed))+
  facet_wrap(~id, scales = "free_y", ncol = 1) +
  xlab("NQT demand (in kwh)")

@@

```{r dotplot-8, fig.cap = "Harmony pairs are shown for all household ids. The darker the colour, the higher the importance of the harmony. Visualizing the pairs in this way helps us to see the important cyclic granularities along the x-axis and facet along with the information that which ones amongst them should be analyzed together." }
elec_sig_split <- elec_harmony_all %>%
  mutate(select_split =  str_split(select_harmony, " ", simplify = TRUE)[,2]) %>% 
  mutate(significant = case_when(
    select_split == "***" ~ "highest",
    select_split == "**" ~ "high",
    select_split == "*" ~ "medium",
    select_split == "" ~ "low"
  )) %>% 
   mutate(rank = case_when(
    select_split == "***" ~ paste(rank, "***", sep = " "),
    select_split == "**" ~  paste(rank, "**", sep = " "),
    select_split == "*" ~  paste(rank, "*", sep = " "),
    select_split == "" ~  paste(rank, "", sep = " ")
  ))

@@

heatplot <- elec_sig_split %>% 
  ggplot() +
  geom_tile(aes(x = x_variable,
                 y = facet_variable, 
                 fill = wpd,
                size = significant)) + 
  scale_fill_continuous(type = "gradient") +
  theme(legend.position = "bottom") +
  facet_wrap(~id, nrow = 2) +
  coord_fixed() + 
  guides(fill = guide_legend()) +
  theme_void() +
  theme_gray(base_size = 12, base_family = "Times") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 60, hjust=1),legend.position = "bottom")
  heatplot
#            
# dotplot <- elec_sig_split %>% 
#   ggplot() +
#   geom_point(aes(x = x_variable,
#                  y = facet_variable, 
#                  size = wpd, 
#                  fill = significant),
#              shape=21,
#              alpha= 0.8) +
#   scale_size_continuous() +
#   theme(legend.position = "bottom") +
#   scale_fill_viridis(discrete = TRUE, direction = 1) +
#   facet_wrap(~id, nrow = 2) +
#   guides(fill = guide_legend(), 
#          size = guide_legend()) +
#   theme_gray(base_size = 12, base_family = "Times") +
#    theme(legend.position = "bottom")
```

@@

elec_rank %>% kable(format = "markdown", caption = "Ranking of the harmonies for the four households are shown. The first four harmonies are unanimously important for all households, after which there is differences in the rankings for most households.")
```

@@

Exploratory data analysis involve many iterations of finding and summarizing patterns. With temporal data available at ever finer scales, exploring periodicity has become overwhelming with so many possible granularities to explore. This work refines the selection of appropriate pairs of granularities by identifying those for which the differences between the displayed distributions is greatest, and rating these selected harmony pairs in order of importance for exploration.

A future direction of work could be to look at more individuals/subjects and group them according to similar periodic behavior. Behaviors across different cyclic granularities would be different for different subjects and one way to find groups would be to actually locate clusters who have similar periodic behavior.

 A common remedy would be to zoom into “interesting” segments, but how do we identify “interesting” segments?

@@

>>>>

---
title: A new metric for automatic discovery of periodic patterns in time series

# to produce blinded version set to 1
blinded: 0

authors:
- name: Sayani Gupta
  thanks: "Email: Sayani.Gupta@monash.edu"
  affiliation: Department of Econometrics and Business Statistics, Monash University, Australia

- name: Rob J Hyndman
  affiliation: Department of Econometrics and Business Statistics, Monash University, Australia

- name: Dianne Cook
  affiliation: Department of Econometrics and Business Statistics, Monash University, Australia

@@

The smart meter data set for eight households in Melbourne has been utilized to see the use of $wpd_{norm}$ proposed in the paper. The data has been cleaned to be a `tsibble` (@wang2019tsibble) containing half-hourly electricity consumption from Jul-2019 to Dec-2019 for each of the households, which is procured by them by downloading their data from the energy supplier/retailer. Demand data for these households are shown in a linear time scale in Figure \ref{fig:linear-scale-8}. It is evident from the range of the demand data that these households vary in consumption levels as well as in their temporal patterns. In the left panel of Figure \ref{fig:linear-scale-8} (a), the linear representation of the entire time period is shown, whereas in the right panel (b) a particular month is shown and furthermore a week has been highlighted to inspect if there is any daily or weekly periodic patterns in their behavior that is reflected when we zoom into the the linear representation of the time series. We have some additional demographics data in terms of the number of members in the household and presence of kids/elderly parents and their profession.

```{r}
demography <- tibble(id = c(1, 2, 3, 4, 5,6, 7, 8), 
       profession = c("academia", "mixed", "industry", "industry", "mixed", "mixed", "academia", "industry"),
       total_members = c(2, 3, 5, 5, 2, 2, 2, 6),
       kids = c("no", "no", "yes", "yes", "no", "no", "no", "yes"),
       old_parents = c("no", "yes", "yes", "yes", "no", "no", "no", "yes"),
       PhD_student = c("no", "yes", "no", "no", "yes", "yes", "yes", "no"))

demography %>% kable(format  = "markdown", caption = "Demographics of the eight households chosen for study")
```

We start the analysis by asking if the ranking of the harmonies make sense for the households, then compare households to get more insights of what these rankings imply and if they could be used to remove some non-interesting harmonies. Furthermore, we see if the display of the significant harmonies could be validated by zooming in the linear representation of the time series.

@@

```{r raw-nqt-8, fig.height = 5, fig.cap="The raw density of the half-hourly demand for the eight households in Panel a. Panel b shows the normal-score-transform half-hourly demand for the same households which has resulted in more symmetric distribution of half-hourly demand."}
elec_raw <- elec %>% 
  ggplot() +
  geom_density(aes(x = kwh)) +
  facet_wrap(~id, scales = "free_y", ncol = 1) +
  xlab("demand (in kwh)") + ggtitle("a") +
  theme(
        strip.text = element_text(size = 10, margin = margin(b = 0, t = 0)))

elec_nqt <- elec %>% 
  dplyr::mutate(kwh_transformed = stats::qqnorm(kwh, plot.it = FALSE)$x) %>% 
  ggplot() +
  geom_density(aes(x = kwh_transformed))+
  facet_wrap(~id, scales = "free_y", ncol = 1) +
  xlab("NQT demand (in kwh)") + ggtitle("b") + theme(
        strip.text = element_text(size = 10, margin = margin(b = 0, t = 0)))

@@

_Selecting and Ranking harmonies for all households_

@@

```{r gravitas-plot-8, message = FALSE, warning=FALSE,fig.cap="Distribution of energy demand shown for household id 1 across hod in x-axis and wd-wnd in facets in a) and just the reverse in b). In c), distribution of energy demand for household id:4 shown across hod and wd-wnd. It can be seen that the differences in distributions are more apparent when viewed in a) as compared to b). It seems like there is more difference in the distributions of hod for b) compared to c). This also confers with the value of the normalised measure shown in Figure 11."}

@@

<!-- From Figure \ref{fig:gravitas-plot-8-id5}, it could further be observed that id5 has only one significant harmony $(hod, dow)$. Apparently, $(hod, wnd/wday)$ which is an important harmony for most households is not important for this one. -->

_Comparing households and validating patterns from linear display_

Figure \ref{fig:dotplot-8} helps to compare between households through the heatmap (b) and validate the results of the heatmap through linear display in (a) . The top panel contains the raw data for a month (Sep-2019). The darker the color of a cell in the heatmap, the more significant a harmony pair it is. Also, the ones with red borders are significant at 95% significance. This plot suggests that there is no significant periodic patterns for id 5, even when the demographics of id 5 is very similar to id 6 and 7. Household id 6 and 7 differ in the sense that for id 6, difference in patterns only during weekday/weekends (two members are in mixed profession), whereas for id 7, all or few other days of the week are also important (members are in academia and hence more flexible routines). The periodic pattern is very similar to id 8, which has a very different demographic property. Both from the linear representation and heatmap, it is clear that id 2 and 3 are similar, both in term of periodic pattern as well as demographics.

```{r gravitas-plot-8-id5, message = FALSE, warning=FALSE, fig.cap="something", eval = FALSE}

@@

@@

```{r dotplot-8, fig.cap = "Harmony pairs are shown for all household ids. The darker the colour, the higher the importance of the harmony. Also, the ones bordered red are selected with 90 percent significance level. Visualizing the pairs in this way helps us to see the important cyclic granularities along the x-axis and facet along with the information that which households should be analyzed together." }
elec_sig_split <- elec_harmony_all %>%
  mutate(select_split =  str_split(select_harmony, " ", simplify = TRUE)[,2]) %>% 
  mutate(significant = case_when(
    select_split == "***" ~ "highest",
    select_split == "**" ~ "high",
    select_split == "*" ~ "medium",
    select_split == "" ~ "low"
  )) %>% 
   mutate(rank = case_when(
    select_split == "***" ~ paste(rank, "***", sep = " "),
    select_split == "**" ~  paste(rank, "**", sep = " "),
    select_split == "*" ~  paste(rank, "*", sep = " "),
    select_split == "" ~  paste(rank, "", sep = " ")
  ))

@@

heatplot <- elec_sig_split %>% 
  mutate(significance_95 = if_else(significant %in% c("high", "highest"), "yes", "no")) %>% 
  ggplot() +
  geom_tile(aes(x = x_variable,
                 y = facet_variable, 
                 fill = wpd, color = significance_95), size = 1) + 
  #scale_fill_gradient(low = "white",high = "steelblue") +
  scale_fill_gradient2() +
  scale_colour_manual(values = c("white","red")) + 
  theme(legend.position = "bottom") +
  facet_wrap(~id, nrow = 1, strip.position = "bottom") +
  coord_fixed() + 
  guides(fill = guide_legend()) +
  theme_void() +
  theme_gray(base_size = 12, base_family = "Times") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 60, hjust=1),legend.position = "bottom") + ggtitle("b")

elec <- read_rds(here("paper/data/elec_all-8.rds")) %>% 
  dplyr::filter(date(reading_datetime) >= ymd("20190701"), date(reading_datetime) < ymd("20191231"), meter_id==1) %>% 
  select(-meter_id) %>% 
  rename("id" = "household_id",
         "date_time" = "reading_datetime")

elec_zoom <-  elec %>%
  as_tibble() %>% 
  dplyr::filter(date(date_time) > as.Date("2019-09-01") & date(date_time) < (as.Date("2019-09-30"))) %>%
  ggplot(aes(x=date_time, y = kwh)) +
  #geom_point(size = 0.1, colour = "black", alpha = 0.3) +
  geom_line(size = 0.1, colour = "blue", alpha = 0.2) +
  facet_wrap(~id, 
             scales = "free_y",
             nrow = 1,
             strip.position =  "top") + xlab("") + 
  theme_void() + theme(strip.text.x = element_blank()) +
  ylab("energy demand in September (kwh)") + ggtitle("a")

elec_zoom/heatplot

#            
# dotplot <- elec_sig_split %>% 
#   ggplot() +
#   geom_point(aes(x = x_variable,
#                  y = facet_variable, 
#                  size = wpd, 
#                  fill = significant),
#              shape=21,
#              alpha= 0.8) +
#   scale_size_continuous() +
#   theme(legend.position = "bottom") +
#   scale_fill_viridis(discrete = TRUE, direction = 1) +
#   facet_wrap(~id, nrow = 2) +
#   guides(fill = guide_legend(), 
#          size = guide_legend()) +
#   theme_gray(base_size = 12, base_family = "Times") +
#    theme(legend.position = "bottom")
```

@@

elec_rank %>% kable(format = "markdown", caption = "Ranking of harmonies for the eight households with significance levels.")
```

@@

Exploratory data analysis involve many iterations of finding and summarizing patterns. With temporal data available at more and more finer scales, exploring periodicity has become overwhelming with so many possible granularities to explore. A common solution would be to zoom into “interesting” segments, but there is no way to know the “interesting” segments a priori. This work refines the search of periodic patterns by identifying those for which the differences between the displayed distributions is greatest, and rating them in order of importance for exploration.

A future direction of work could be to look at more individuals/subjects and group them according to similar periodic behavior. Behaviors across different harmonies would be varying for subjects and it would be hard to track the behavior when the number of individuals rise. One way to find groups would be to actually locate clusters who have similar periodic behavior.

@@

# Bibliography

```{r write-bib, eval = FALSE}
write_bib(c("rmarkdown", "bookdown", "rticles", "knitr", "ggplot2", "dplyr", "tsibble", "gravitas", "ggpubr", "kableExtra", "magrittr", "readr", "lubridate", "renv", "archive", "hakear", "patchwork", "ggpubr"), file = "rpkgs.bib")
```
